Evolution of Processes in Delivering Reliable Software in the European Scientific Arena

### 2018-06-11 paper-ieee-trans.tex

In the recent decades the development of new technologies in the distributed computing field (i.e. Grid and Cloud computing) have contributed to
providing new tools and services able to match the scientific community requirements. In Europe various projects have been designed in order to develop and maintain software able to support and manage computing infrastructure. In this context, activities have been performed to deliver reliable software in the European research arena by using innovative software engineering solutions.

This work examines the evolution of the software engineering methodologies used in past and
current European Commission--funded projects to illustrate how the research on software
reliability has progressed over the last 15 years since the
first project in 2001. We summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. 
Recent software development projects show good evidences of such insights, where the enforcement of quality assurance
procedures is present since the very early stages of the software development process.

@@

Major developments carried throughout the years where focused on providing new
functionalities, while in the early days, the adoption of software engineering methodologies had a lower priority.
The released software was less tested, resulting in a higher rate of
rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.
% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
It became then apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, incoming projects progressively
devoted a substantial part of the software lifecycle to the improvement of the quality of the
software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

Analysing the EC--funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization) motivated this trend. All of which
are complemented by the emergence of automation and event--response tools.

@@

\section{Background}
\label{sec:b}

\section{Research methodology}
\label{sec:rm}

@@

>>>>

From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.

@@

Major developments carried throughout the years where focused on providing new
functionalities, while in the early days, the adoption of software engineering methodologies had a lower priority.
The released software was less tested, resulting in a higher rate of
rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities \cite{aiftimiei}.
% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
It became then apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, incoming projects progressively
devoted a substantial part of the software lifecycle to the improvement of the quality of the
software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

Analysing the EC--funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures \cite{aiftimiei}. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend. All of which
are complemented by the emergence of automation and event--response tools.

@@

@@

\bibitem{soltesz}
S. Soltesz, and H. P\"{o}tzl, et. al.,
Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
\emph{SIGOPS Oper. Syst. Rev.},
vol. 41, n. 3, pp. 275--287, 2007.
